\MakeUppercase{Layered $H$-Partitions of $k$-Planar Graphs} \thanks{This work was partly funded by NSERC and MRI.}

### 2019-05-12 k-planar.tex
Polishing...

\begin{abstract}
  Layered $H$-partitions of graphs (with small layered width, in which $H$ has small treewidth) are a recently-introduced tool that have been used to solve longstanding problems on queue layouts, non-repetitive colouring, and 3-d graph drawing.  Such partitions are known to exist for planar graphs and, more generally, bounded genus graphs.  In the current paper, we prove that every $k$-planar graph has a layered $H$-partition of layered width $O(k)$ in which $H$ has treewidth $O(c^k)$, for some constant $c$. This is the first result of this type of a non-minor closed class of graphs and implies that $k$-planar graphs have both queue number and non-repetitive chromatic number upper-bounded by a function of $k$.  The former result was previously shown using a combination of $H$-partitions for planar graphs with \textit{ad hoc} methods. The latter result is new, for all $k\ge 3$.
\end{abstract}

@@

A \emph{layering} of a graph $G$ is a sequence $\mathcal{L}=\langle V_0,V_1,\ldots,V_h\rangle$ such that $\{V_0,V_1,\ldots,V_h\}$ is a partition of $V(G)$ and there is no edge $vw\in E(G)$ such that $v\in V_i$, $w\in V_j$ and $|j-i|>1$.  For any partition $\mathcal{P}=\{S_1,\ldots,S_p$ of $V(G)\}$, a \emph{quotient graph} $H=G/\mathcal{P}$ has a $p$-element vertex set $V(H)=\{x_1,\ldots,x_p\}$ and $x_ix_j\in E(H)$ if and only there exists an edge $vw\in E(G)$ such that $v\in S_i$ and $w\in S_j$. To highlight the importance of the quotient graph $H$, we call $\mathcal{P}$ an $H$-partition and write this concisely as $\mathcal{P}=\{S_x : x\in V(H)\}$ so that each element of $\mathcal{P}$ is indexed by the vertex it creates in $H$.\footnote{An alternative definition of a layering of $G$ is a layered $H$-partition of $G$ where $H$ is a path.}  A layered $H$-partition $(\mathcal{L},\mathcal{P})$ of a graph $G$ consists of a layering $\mathcal{L}$ of $G$ and an $H$-partition $\mathcal{P}$ of $G$. The \emph{layered width} of $(\mathcal{H},\mathcal{P})$ is $\max\{|L\cap P|: L\in\mathcal{L},\, P\in\mathcal{P}\}$.

Layered $H$-partitions of small layered width in which $H$ has some additional property are a useful tool for a number of graph problems. \citet{dujmovic.joret.ea:planar} show that, if each graph $G$ in some family $\mathcal{G}$ of graphs has a layered $H$-partition for which the layered-width of the partition and the treewidth of $H$ are each upper-bounded by some constant $c_\mathcal{G}$, then each of the following quantities are upper bounded by a constant, for every $G\in\mathcal{G}$: queue number \cite{X}, track number \cite{X}, etc \cite{X}, etc \cite{X}, etc \cite{X}. \citet{dujmovic.esperet.ea:planar} add non-repetitive chromatic number to this list.  \citet{dujmovic.joret.ea:planar} show that two common families of graphs have such $H$-partitions: planar graphs and (more generally) bounded-genus graphs.

@@

Under these definitions $0$-planar graphs are exactly planar graphs and $0$-plane graphs are plane graphs. Thus, like genus-$g$ graphs, $k$-planar graphs are a generalization of planar graphs (which are genus-$0$ graphs).  Unlike genus-$g$ graphs, however, the family of $k$-planar graphs is not minor closed. A graph $G'$ obtained from a $k$-planar graph $G$ by edge deletions and edge contractions may or may not be $k$-planar.

We prove the following result:
\begin{thm}\thmlabel{k-planar}
  Every $k$-planar graph has a layered $H$-partition of layered-width at most $24k^2 + 36k + 12$ in which $H$ has treewidth at most $(3^{k+2}-3)/2$.
\end{thm}

@@

The purpose of this section is to prove \thmref{k-planar}.  In this section, all graphs are finite, simple, an undirected.  For any graph $G$ and any set $S$ (typically $S\subseteq V(G)$) we use $G[S]$ to denote the graph with vertex set $V(G)\cap S$ and edge set $\{uv\in E(G) : u,v\in S\}$.  We use $G-S$ as a shorthand for $G[V(G)\setminus S]$ and for a graph $G'$, we use $G-G'$ as a shorthand for $G-V(G')$.

@@

Let $T$ be a rooted tree rooted at $r\in V(T)$. The \emph{$T$-depth} of a node $x\in V(T)$ is equal to the length of the path from $x$ to $r$ in $T$. A $P$ path in $T$ is \emph{vertical} if, for each integer $i$, $P$ contains at most one vertex of $T$-depth $i$.  The deepest vertex in a vertical path $P$ is called  $P$'s \emph{lower endpoint} and the other (shallowest) vertex in $P$ is called $P$'s \emph{upper endpoint}. For any node $v\in T$, we call each node on the path, in $T$, from $v$ to $r$ a \emph{$T$-ancestor} of $v$.

@@

For the remainder of this section, $\Delta$ is a triangulation, and $T$ is a BFS spanning tree of $\Delta$ rooted at some degree-3 vertex $r$ on the outer face of $\Delta$.

@@

  \item An inner face $\tau_x$ of $N_x$ that determines a closed tripod $\overline{Y}_x$ and an open tripod $Y_x$ in $N_x$.

@@

  \item \proplabel{ancestor-boundary} For each non-root node $x\in V(T)$, there exists a partition of $V(F_x)$ into $g\le 3$ sets $P_1,\ldots,P_g$ such that, for each $i\in\{1,\ldots,g\}$, $F[P_i]$ is a path that is contained in $Y_a$ for some $K$-ancestor $a$ of $x$ \emph{or} $F[P_i]$ is a path or cycle in $F_{r_K}$.

  \item \proplabel{partition} $\{Y_x : x\in V(K)\}$ is a partition of $V(\Delta)\setminus\{r\}\setminus V(F_{r_K})$.
\end{enumerate}

@@

Let $G$ be a $k$-plane graph.  We will assume, for ease of exposition, that any point $p\in\R^2$ is involved in at most one crossing $(p,vw,xy)$ of $G$. This assumption is not critical since it can be enforced by a slight deformation of the edges of $G$.

Let $G^+$ be the graph obtained as follows:
\begin{compactenum}
  \item Add a dummy vertex at each crossing in $G$ to obtain a plane graph and then add edges to this graph to obtain a triangulation $G^+$.
  \item Add a complete graph $K_4$ that contains $G^{+}$ in one of its inner faces, $F_0$, and then add edges to the resulting graph to obtain a triangulation $G^{++}$.
\end{compactenum}
Let $r$ be the vertex of the $K_4$ not incident on $F_0$ and let $T$ be a BFS spanning tree of $G^+$ rooted at $r$. Observe that, for any edge $vw\in E(G)$, $\dist_{G^+}(v,w) \le k+1$ and therefore, for any two vertices $v,w\in V(G)$,
\begin{equation} 
  \dist_{G^+}(v,w) \le (k+1)\cdot \dist_G(v,w) \eqlabel{distance-preserving}
\end{equation}
For each integer $i\ge 0$, let $V_i=\{v\in V(G^+): \dist_{T}(r,v)=i\}$ and observe that $\{V_0,V_1,\ldots,V_{h}\}$ is a layering of $G^+$ (but not of $G$).

@@

For each vertex $v\in V(G^+)$, let $P_v=\{a\in V(K): v\in V(N_a-F_a)\}$.  
By \propref{all-in}, $P_{v}$ contains the root $r_K$ of $K$.  By \propref{children}, $K[P_v]$ is a vertical path (whose lower endpoint is the unique node $x\in V(K)$ such that $v\in Y_x$). Therefore $K[P_v]$ is a vertical path in $K$ that contains $r_K$.  We are about to repeatedly use the fact that the common intersection of two or more vertical paths is a vertical path.

For each edge $vw\in E(G)$, consider the walk $W_{vw}$ in $G^+$ that corresponds to the sequence of vertices in $G^+$ that we encounter while walking along the curve $vw$ from $v$ to $w$. We define $A_{vw}=\bigcap_{z\in W_{vw}} P_z$.  Since $W_{vw}\subseteq V(G^+)$, $A_{vw}$ is well-defined.  For each $z\in W_{vw}$, $K[P_{z}]$ is a vertical path in $K$ that contains $r_K$.   Therefore $A_{vw}$ is a non-empty vertical path in $K$.  For any vertex $v\in V(G)$ define $A_v=\bigcap_{vw\in E(G)} A_{vw}$.  Again, $K[A_v]$ is a non-empty vertical path in $K$ that contains $r_K$.  We define $a(v)$ to be the lower endpoint of $K[A_v]$.  For every node $x\in V(K)$, define the \emph{separator}
\[
   S_x = \{v\in V(G): a(v)=x \} \enspace .
\]
Clearly $\mathcal{S}=\{S_x:x\in V(K)\}$ is a partition of $V(G)$.  We will use $\mathcal{S}$ as the partition that defines our quotient graph $H=G/\mathcal{S}$ with vertex set $V(H)=V(K)$ and the edge $xy\in E(H)$ if and only if there exists an edge $vw\in E(G)$ with $v\in S_x$ and $w\in S_y$.  First we argue that this partition has small width with respect to the layering defined by $T$ (recall that this is a layering of $G^+$, not a layering of $G$).

@@

\begin{proof}
  Since $v\in S_x$, the lower endpoint of $K[A_{vw}]$ is $x$ for some edge $vw\in E(G)$.  This implies that $V(W_{vw})\subset V(N_x-F_x)$ but $V(W_{vw}) \not\subseteq V(N_{x'}-F_{x'})$ for any any child $x'$ of $x$ in $K$.  This implies that $W_{vw}$ contains a vertex $v'\in Y_x$.  Since $G$ is $k$-planar and $v'\in V_{vw}$ this implies that $\dist_{G^+}(v,v') \le k+1$.
  By the triangle inequality,     
  \[
    \dist_{G^+}(r,v') \le \dist_{G^+}(r,v) + \dist_{G^+}(v,v') \le \dist_{G^+}(r,v) + k + 1 \enspace .
  \]
  Similarly, $\dist_{G^+}(r,v) \le \dist_{G^+}(r, v') + k+1$.  Therefore
  $|\dist_{G^+}(r,v)-\dist_{G^+}(r,v')| \le k+1$ which establishes the claim, since $T$ is BFS tree, so $\dist_{T}(r,z)=\dist_{G^+}(r,z)$ for all $z\in V(G^+)$.
\end{proof}

@@

Next we describe a tree decomposition $(B_x:x\in V(K))$ of $H$.  We deliberately use the same node set $V(H)=V(K)$ here, because the subtree $K[x]:=K[\{z:x\in B_z\}]$ will be rooted at $x$.

@@

  Consider the graph $A$ that contains the edges and vertices of $F_{a}$ for all $K$-ancestors $a$ of $x$. See \figref{A}.  It is helpful to think of this graph as follows:  We begin with $F_{a_0}$ where $a_0=r_K$ is the root of $K$.  We then cut $F_{a_0}$ into two pieces $F^{\bar{x}}_{a_0}$ and $F^x_{a_0}$ where $F^x_{a_0}$ is the piece that contains $F_x$. (The path that makes this cut contains two legs $Q_i$ and $Q_j$ of a tripod $Y$ and one edge $v_iv_j$ of its central triangle $\tau$.)  This process continues by partitioning $F_{a_1}:=F^x_{a_0}$ recursively.  The process terminates after some number $d$ of iterations, when $F^x_{a_{d-1}}=F_x=F_{d}$.

@@

  However, $w\in S_a$, so there exists an edge $ww'\in E(G)$ such that   $W_{ww'}$ is contained in $V(W_{ww'})\subseteq V(N_{a_i}-F_{a_i})$ but $V(W_{ww'}\not\subseteq V(N_{a_{i+1}}-F_{a_{i+1}})$.  Therefore $W_{ww'}$ must contain a first vertex $w^*\in V(F_{a_{i+1}}-F_{a_{i}})$.

  Summarizing, we have a walk $w'=v_0,\ldots,v_\ell=w^*$ in $G^+$ such that
  \begin{compactenum}[(i)]
    \item $v_0\in V(N_x-F_x)$;

@@

    \item $\ell\le k+1$.
  \end{compactenum}
  We call any walk with properties (i)--(iv) an \emph{$i$-walk}. Therefore, to upper-bound $|B_x|$, it suffices to upper-bound the number of integers $i\in\{0,\ldots,d-1\}$ such that there exists an $i$-walk.

  For each $v'\in V(G^+)$, 
  \[  
      d_{v'}=\max\{j\in\{0,\ldots,d\} : v'\in V(N_{a_j}-F_{a_j})
  \]
  (Intuitively, $d_{v'}$ is the $K$-depth of the lower endpoint of $K[P_{v'}]$ or the $K$-depth ($d$) of $x$, whichever is shallower.)

  Let $v_0,\ldots,v_\ell$ be an $i$-walk and and consider the subsequence $d_0,\ldots,d_{\ell'}$ of $d_{v_0},\ldots,d_{v_\ell}$ consisting of only $d_{v_0}$ and those (record-breaking) $d_{v_i}$ such that $d_{v_i}<\min\{d_{v_0},\ldots,d_{v_{i-1}}\}$.  Observe that $d_0=d$ and $d_{\ell'}=i$.

  By \propref{ancestor-boundary}, for every $j\in\{1,\ldots,d\}$, there is a set $X_{j}$, of at most 3 integers $\alpha < j$ such that such that every vertex $q\in V(F_{a_j})$ is in $V(Y_{a_{\alpha}})$ for some $\alpha\in X_j$.
  Assign a distinct colour $c(j,\alpha)\in\{1,2,3\}$ for each $\alpha\in X_j$ and do this assignment for each $j\in\{1,\ldots,d\}$.  By definition, for any $j\in\{1,\ldots,d\}$ and any colour $\gamma\in\{1,2,3\}$, there is at most one $\alpha=\alpha(j,\gamma)$ such that $c(j,\alpha)=\gamma$.

@@

  The proof is now almost complete.  Any $i$-walk defines a colour sequence $c_0,\ldots,c_{\ell'}$.  If some $i'$-walk defines the same colour sequence $c_0,\ldots,c_{\ell'}$ then
  \[
        i = (\alpha(\alpha(\alpha(d,c_0),c_1),c_2)\cdots),c_{\ell'-1}) = i' \enspace ,
  \]
  so $i=i'$.  Since there are at most $3^{\ell'}$ colour sequences of length $\ell'$ and $\ell'\le \ell \le k+1$, the number of $i\in\{0,\ldots,d-1\}$ such that there exists an $i$-walk is at most
  \[
     |B_x| \le \sum_{\ell=1}^{k+1} 3^\ell = (3^{k+2}-3)/2 \eqlabel{bag-size} \enspace . \qedhere
  \]
\end{proof}

We now have all the parts needed for the proof of \thmref{k-planar}.

@@

  By \clmref{bag-size}, $\mathcal{S}$ produces a graph $H=G/\mathcal{S}$ of treewidth $(3^{k+2}-1)/2$.  By \clmref{width-of-g-plus}, for each $x\in V(H)$ and each integer $i$, $|B_x\cap V_i|\le 24k+12$, so $|B_x\cap V_i'|\le (k+1)(24k + 12) = 24k^2+36k+12$. 
\end{proof}

>>>>

\begin{abstract}
  Layered $H$-partitions of graphs (with small layered width, in which $H$ has small treewidth) are a recently-introduced tool that have been used to solve longstanding problems on queue layouts, non-repetitive colouring, and 3-d graph drawing.  Such partitions are known to exist for planar graphs and, more generally, bounded genus graphs.  In the current paper, we prove that every $k$-planar graph has a layered $H$-partition of layered width $O(k^2)$ in which $H$ has treewidth $O(c^k)$, for some constant $c$. This is the first result of this type for a non-minor-closed class of graphs and implies that $k$-planar graphs have both queue number and non-repetitive chromatic number upper-bounded by a function of $k$.  The former result was previously shown using a combination of $H$-partitions for planar graphs with \textit{ad hoc} methods. The latter result is new, for all $k\ge 1$.
\end{abstract}

@@

A \emph{layering} of a graph $G$ is a sequence $\mathcal{L}=\langle V_0,V_1,\ldots,V_h\rangle$ such that $\{V_0,V_1,\ldots,V_h\}$ is a partition of $V(G)$ and there is no edge $vw\in E(G)$ such that $v\in V_i$, $w\in V_j$ and $|j-i|>1$.  For any partition $\mathcal{P}=\{S_1,\ldots,S_p\}$ of $V(G)$, a \emph{quotient graph} $H=G/\mathcal{P}$ has a $p$-element vertex set $V(H)=\{x_1,\ldots,x_p\}$ and $x_ix_j\in E(H)$ if and only there exists an edge $vw\in E(G)$ such that $v\in S_i$ and $w\in S_j$. To highlight the importance of the quotient graph $H$, we call $\mathcal{P}$ an $H$-partition and write this concisely as $\mathcal{P}=\{S_x : x\in V(H)\}$ so that each element of $\mathcal{P}$ is indexed by the vertex it creates in $H$.\footnote{An alternative definition of a layering of $G$ is a layered $H$-partition of $G$ where $H$ is a path.}  A layered $H$-partition $(\mathcal{L},\mathcal{P})$ of a graph $G$ consists of a layering $\mathcal{L}$ of $G$ and an $H$-partition $\mathcal{P}$ of $G$. The \emph{layered width} of $(\mathcal{H},\mathcal{P})$ is $\max\{|L\cap P|: L\in\mathcal{L},\, P\in\mathcal{P}\}$.

Layered $H$-partitions of small layered width in which $H$ has some additional property are a useful tool for a number of graph problems. \citet{dujmovic.joret.ea:planar} show that, if each graph $G$ in some family $\mathcal{G}$ of graphs has a layered $H$-partition for which the layered-width of the partition and the treewidth of $H$ are each upper-bounded by some constant $c_\mathcal{G}$, then each of the following quantities are upper bounded by a constant, for every $G\in\mathcal{G}$: queue number \cite{X}, track number \cite{X}, layered treewidth \cite{X}, etc \cite{X}, etc \cite{X}, etc \cite{X}. \citet{dujmovic.esperet.ea:planar} add non-repetitive chromatic number to this list.  \citet{dujmovic.joret.ea:planar} show that two common families of graphs have such $H$-partitions: planar graphs and (more generally) bounded-genus graphs.

@@

Under these definitions $0$-planar graphs are exactly planar graphs and $0$-plane graphs are exactly plane graphs. Thus, like genus-$g$ graphs, $k$-planar graphs are a generalization of planar graphs (which are genus-$0$ graphs).  Unlike genus-$g$ graphs, however, the family of $k$-planar graphs is not minor closed. A graph $G'$ obtained from a $k$-planar graph $G$ by edge deletions and edge contractions may or may not be $k$-planar.

We prove the following result:
\begin{thm}\thmlabel{k-planar}
  Every $k$-planar graph has a layered $H$-partition of layered-width at most $24k^2 + 36k + 12$ in which $H$ has treewidth at most $(3^{k+2}-5)/2$.
\end{thm}

@@

The purpose of this section is to prove \thmref{k-planar}.  In this section, all graphs are finite, simple, an undirected.  For any graph $G$ and any set $S$ (typically $S\subseteq V(G)$) we use $G[S]$ to denote the graph with vertex set $V(G)\cap S$ and edge set $\{uv\in E(G) : u,v\in S\}$.  We use $G-S$ as a shorthand for $G[V(G)\setminus S]$ and for a graph $G'$, we use $G-G'$ as a shorthand for $G-V(G')$.  We use $G'\subseteq G$ to denote subgraph containment, i.e., $V(G')\subseteq V(G)$ and $E(G')\subseteq E(G)$.

@@

Let $T$ be a rooted tree rooted at some node $r\in V(T)$. The \emph{$T$-depth} of a node $x\in V(T)$ is equal to the length of the path from $x$ to $r$ in $T$. A $P$ path in $T$ is \emph{vertical} if, for each integer $i$, $P$ contains at most one vertex of $T$-depth $i$.  The deepest vertex in a vertical path $P$ is called  $P$'s \emph{lower endpoint} and the shallowest vertex in $P$ is called $P$'s \emph{upper endpoint}. For any node $v\in T$, we call each node on the path, in $T$, from $v$ to $r$ a \emph{$T$-ancestor} of $v$.

@@

For the remainder of this section, $\Delta$ is a triangulation, $r$ is a degree-3 vertex on the outer face of $\Delta$, and $T$ is a BFS spanning tree of $\Delta$ rooted at $r$.

@@

  \item An inner face $\tau_x$ of $N_x$ that determines a closed tripod $\overline{Y}_x$ and an open tripod $Y_x=\overline{Y}_x-F_x$ in $N_x$.

@@

  \item \proplabel{ancestor-boundary} For each non-root node $x\in V(T)$, there exists a partition of $V(F_x)$ into $g\le 3$ sets $P_1,\ldots,P_g$ such that, for each $i\in\{1,\ldots,g\}$, either:
  \begin{compactenum}
    \item $F[P_i]$ is a path that is contained in $Y_a$ for some $K$-ancestor $a$ of $x$;
    \item $F[P_i]$ is a cycle and there exists an edge of $P_i$ whose removal produces a path that is contained in $Y_a$ for some $K$-ancestor $a$ of $x$; or 
    \item $F[P_i]$ is a path or cycle in $F_{r_K}$.
  \end{compactenum}
  \item \proplabel{partition} $\{Y_x : x\in V(K)\}$ is a partition of $V(\Delta)\setminus\{r\}\setminus V(F_{r_K})$.
\end{enumerate}

@@

Let $G$ be a $k$-plane graph.  We will assume, for ease of exposition, that any point $p\in\R^2$ is involved in at most one crossing $(p,vw,xy)$ of $G$. This assumption is not critical since it can be enforced by a slight deformation of the edges of $G$.   Consider the following plane graphs:
\begin{compactenum}
  \item Add a dummy vertex at each crossing in $G$ to obtain a plane graph and then add edges to this graph to obtain a triangulation $G^+$.
  \item Add a complete graph $K_4$ that contains $G^{+}$ in one of its inner faces, $F_0$, and then add edges to the resulting graph to obtain a triangulation $G^{++}$.
\end{compactenum}
Let $r$ be the vertex of the $K_4$ not incident on $F_0$ and let $T$ be a BFS spanning tree of $G^{++}$ rooted at $r$. Observe that, for any edge $vw\in E(G)$, 
\[  \dist_{G^+}(v,w) \le k+1 \enspace . \]
%  and therefore, for any two vertices $v,w\in V(G)$,
% \begin{equation} 
%   \dist_{G^+}(v,w) \le (k+1)\cdot \dist_G(v,w) \eqlabel{distance-preserving}
% \end{equation}
For each integer $i\ge 0$, let $V_i=\{v\in V(G^+): \dist_{T}(r,v)=i\}$ and observe that $\{V_0,V_1,\ldots,V_{h}\}$ is a layering of $G^+$ (but not of $G$).

@@

For each vertex $z\in V(G^+)$, let $P_z=\{a\in V(K): z\in V(N_a-F_a)\}$.  
By \propref{all-in}, $P_{z}$ contains the root $r_K$ of $K$.  By \propref{children}, $K[P_z]$ is a vertical path (whose lower endpoint is the unique node $x\in V(K)$ such that $z\in Y_x$). Therefore $K[P_z]$ is a vertical path in $K$ that contains $r_K$.  We are about to repeatedly use the fact that the common intersection of two or more vertical paths is a vertical path.

For each edge $vw\in E(G)$, consider the walk $W_{vw}$ in $G^+$ that corresponds to the sequence of vertices in $G^+$ encountered while walking along the curve $vw$ from $v$ to $w$. We define $A_{vw}=\bigcap_{z\in W_{vw}} P_z$.  Since $V(W_{vw})\subseteq V(G^+)$, $A_{vw}$ is well-defined.  For each $z\in V(W_{vw})$, $K[P_{z}]$ is a vertical path in $K$ that contains $r_K$.   Therefore $A_{vw}$ is a non-empty vertical path in $K$.  For any vertex $v\in V(G)$ define $A_v=\bigcap_{vw\in E(G)} A_{vw}$.  Again, $K[A_v]$ is a non-empty vertical path in $K$ that contains $r_K$.  We define $a(v)$ to be the lower endpoint of $K[A_v]$.  For every node $x\in V(K)$, define the \emph{separator}
\[
   S_x = \{v\in V(G): a(v)=x \} \enspace .
\]
Clearly $\mathcal{S}=\{S_x:x\in V(K)\}$ is a partition of $V(G)$.  We will use $\mathcal{S}$ as the partition that defines our quotient graph $H=G/\mathcal{S}$ with vertex set $V(H)=V(K)$ and the edge $xy\in E(H)$ if and only if there exists an edge $vw\in E(G)$ with $v\in S_x$ and $w\in S_y$.  First we argue that this partition has small width with respect to the layering $\{V_0,\ldots,V_h\}$ defined by $T$ (recall that this is a layering of $G^+$, not a layering of $G$).

@@

\begin{proof}
  Since $v\in S_x$, the lower endpoint of $K[A_{vw}]$ is $x$ for some edge $vw\in E(G)$.  This implies that $W_{vw}\subset N_x-F_x$ but $W_{vw} \not\subseteq N_{x'}-F_{x'}$ for any any child $x'$ of $x$ in $K$. Therefore, \propref{children} and \propref{ancestor-boundary} imply that $W_{vw}$ contains a vertex $v'\in Y_x$.  Since $G$ is $k$-planar and $v'\in V(W_{vw})$ this implies that $\dist_{G^+}(v,v') \le k+1$.
  By the triangle inequality,     
  \[
    \dist_{G^+}(r,v') \le \dist_{G^+}(r,v) + \dist_{G^+}(v,v') \le \dist_{G^+}(r,v) + k + 1 \enspace .
  \]
  Similarly, $\dist_{G^+}(r,v) \le \dist_{G^+}(r, v') + k+1$.  Therefore
  $|\dist_{G^+}(r,v)-\dist_{G^+}(r,v')| \le k+1$ which establishes the claim, since $T$ is BFS tree rooted at $r$, so $\dist_{T}(r,z)=\dist_{G^+}(r,z)$ for all $z\in V(G^+)$.
\end{proof}

@@

Next we describe a tree decomposition $\{B_x:x\in V(K)\}$ of $H$.  We deliberately use the same node set $V(H)=V(K)$ here, because the subtree $K[x]:=K[\{z:x\in B_z\}]$ will be rooted at $x$.

@@

  Consider the graph $A$ that contains the edges and vertices of $F_{a}$ for all $K$-ancestors $a$ of $x$. See \figref{A}.  It is helpful to think of this graph as follows:  We begin with $F_{a_0}$ where $a_0=r_K$ is the root of $K$.  We then cut $F_{a_0}$ into two pieces $F^{\bar{x}}_{a_0}$ and $F^x_{a_0}$ where $F^x_{a_0}$ is the piece that contains $F_x$. (The path that makes this cut contains two legs $Q_i$ and $Q_j$ of a tripod $Y$ and one edge $v_iv_j$ of its central triangle $\tau$.)  This process continues by partitioning $F_{a_1}:=F^x_{a_0}$ recursively.  The process terminates after some number $d$ of iterations, when $F^x_{a_{d-1}}=F_x=F_{a_d}$.

@@

  However, $w\in S_{_i}$, so there exists an edge $ww'\in E(G)$ such that   $W_{ww'}$ is contained in $W_{ww'}\subseteq N_{a_i}-F_{a_i}$ but $W_{ww'}\not\subseteq N_{a_{i+1}}-F_{a_{i+1}}$.  Therefore $W_{ww'}$ must contain a first vertex $w^*\in V(F_{a_{i+1}}-F_{a_{i}})\subseteq Y_{a_i}$.

  Summarizing, we have a walk $w'=v_0,v_1\ldots,v_{\ell-1},v_\ell=w^*$ in $G^+$ such that
  \begin{compactenum}[(i)]
    \item $v_0\in V(N_x-F_x)$;

@@

    \item $\ell\le k+1$.
  \end{compactenum}
  We call any walk with properties (i)--(iv) an \emph{$i$-walk}. Therefore, to upper-bound $|B_x|$, it suffices to upper-bound the number of $i\in\{0,\ldots,d-1\}$ such that there exists an $i$-walk.

  For each $v'\in V(G^+)$, define
  \[  
      d_{v'}=\max\{j\in\{0,\ldots,d\} : v'\in V(N_{a_j}-F_{a_j})
  \]
  (Intuitively, $d_{v'}$ is the $K$-depth of the lower endpoint of $K[P_{v'}]$ or the $K$-depth ($d$) of $x$, whichever is shallower.)

  Let $v_0,\ldots,v_\ell$ be an $i$-walk and consider the subsequence $d_0,\ldots,d_{\ell'}$ of $d_{v_0},\ldots,d_{v_\ell}$ consisting of only $d_{v_0}$ and those (record-breaking) $d_{v_j}$ such that $d_{v_j}<\min\{d_{v_0},\ldots,d_{v_{j-1}}\}$.  Observe that $d_0=d$ and $d_{\ell'}=i$.

  By \propref{ancestor-boundary}, for every $j\in\{1,\ldots,d\}$, there is a set $X_{j}$, of at most 3 integers $\alpha < j$ such that every vertex $q\in V(F_{a_j})$ is in $V(Y_{a_{\alpha}})$ for some $\alpha\in X_j$.
  Assign a distinct colour $c(j,\alpha)\in\{1,2,3\}$ for each $\alpha\in X_j$ and do this assignment for each $j\in\{1,\ldots,d\}$.  By definition, for any $j\in\{1,\ldots,d\}$ and any colour $\gamma\in\{1,2,3\}$, there is at most one $\alpha=\alpha(j,\gamma)$ such that $c(j,\alpha)=\gamma$.

@@

  The proof is now almost complete.  Any $i$-walk defines a colour sequence $c_0,\ldots,c_{\ell'}$.  If some $i'$-walk defines the same colour sequence $c_0,\ldots,c_{\ell'}$ then
  \[
        i = (\alpha(\alpha(\alpha(d,c_0),c_1),c_2)\cdots),c_{\ell'-1}) = i' \enspace ,
  \]
  so $i=i'$.  Since there are at most $3^{\ell'}$ colour sequences of length $\ell'$ and $\ell'\le \ell \le k+1$, the number of $i\in\{0,\ldots,d-1\}$ such that there exists an $i$-walk is at most
  \[
     \sum_{\ell=1}^{k+1} 3^\ell = (3^{k+2}-3)/2 \ge |B_x| \eqlabel{bag-size} \enspace . \qedhere
  \]
\end{proof}

\note{PM}{I know the analysis in \clmref{bag-width} is not tight---the base $3$ of the exponent can be reduced.  Is it even exponential though?}
We now have all the parts needed for the proof of \thmref{k-planar}.

@@

  By \clmref{bag-size}, $\mathcal{S}$ produces a graph $H=G/\mathcal{S}$ of treewidth $(3^{k+2}-5)/2$.  By \clmref{width-of-g-plus}, for each $x\in V(H)$ and each integer $i$, $|B_x\cap V_i|\le 24k+12$, so $|B_x\cap V_i'|\le (k+1)(24k + 12) = 24k^2+36k+12$. 
\end{proof}
